---
title: "CAPDAP_exploration"
author: "ML"
format: html
editor: visual
---

## CAPDAP Data Exploration

preparing my data and running data analysis

```{r}
rm (list = ls())
library(here)
library(tidyverse)
library(multcomp)
library(ggfortify)
```

```{r}
WC <- read.csv(here("data", "CAPDAP_CleanLog_Final copy.csv"))
```

## a. Check your data for mistakes and outliers. If there are outliers, decide what you will do with them and provide a justification for your choice and then take whatever action you need to deal with them.

lets look at some histograms of pH, DO and Chl data to see the overall frequency of values

```{r}
ggplot(WC,
  aes(pH))+
    geom_histogram()+
    theme_bw()
```

examining this, there is a left distribution of the data, centered around 7.1/7.2 and an outlier above 7.6 there is one really high point beyond 7.6 pH - i'll investigate

```{r}
View(head(WC))
```

conductivity and DO values taken at this time also seem odd - this likely could have been cause by an issue with the hydrolab so I'm going to take this point out.

```{r}
WC <- WC %>%
  filter(pH<7.5)
```

this got rid of one observation

let's check again

```{r}
ggplot(WC,
  aes(pH))+
    geom_histogram()+
    theme_bw()
```

no further concerns with pH

lets look at the distribution of DO data

```{r}
ggplot(WC,
  aes(Dissolved_Oxygen_Saturation))+
    geom_histogram()+
    theme_bw()
```

looks like there are two peaks in the data around 55 and 70, some one-off values in the high 70s, but no dramatic gaps in the data

now lets look at the Chlorophyll data

```{r}
ggplot(WC,
  aes(Chlorophyll))+
    geom_histogram()+
    theme_bw()
```

one large outlier - likely an issue with a lab sensor, so I'm going to remove the point

```{r}
WC <- WC %>%
  filter(Chlorophyll<10)
```

this got rid of one more observation

let's check again

```{r}
ggplot(WC,
  aes(Chlorophyll))+
    geom_histogram()+
    theme_bw()
```

This data set has a more normal distribution of data, with a few spare points in the range above 4. However, I do think that these are genuinely recorded data points despite them being outliers.

now, I'll plot anticipated results of my analysis

```{r}
ggplot(WC,
  aes(x=Chlorophyll, y=Dissolved_Oxygen_Saturation))+
    geom_point()+
    theme_bw()
```

There are no more wild outliers of concern here. I won't start any analysis, but it is interesting to see these variables plotted against one another.

to plot time, I'm going to need to change my data types, so I'll do all of that now while i'm here

fix data types

```{r}
WC$Site_Type <- as.factor(WC$Site_Type)
WC$Date<-mdy(WC$Date)
WC$Time<-parse_time(WC$Time, '%H:%M:%S %p')
WC$DateTime <- with(WC, Date + Time)

```

The time values are only going up to 12:59, but I want them to written in military time. Recruiting Erika's help here.

```{r}
ggplot(WC,aes(x=Time, y=pH))+
    geom_point()+
    scale_x_time()+
    theme_bw()
```

Again, there are no more wild outliers of concern here and I won't start any analysis, but it is interesting to see these variables plotted against one another, and this reminds me that I have an issue with my time scale!

check levels of factor data

```{r}
levels(WC$Site_Type)
```

no issues here so no need to alter factor data

## b.     Change any of the variable names or create new variables that you will eventually need for your data analysis.

I'm going to add one new variable to use in my data analysis

use mutate () to create a new column that has the complete time stamp for each observation called "MDY_HMS"

```{r}
WC<-WC%>%
  mutate(
   MDY_HMS = c(WC$Date+WC$Time)
  )
```

change variable names

```{r}
WC<-WC%>%
  mutate(
   Temperature = WC$Temperature_C,
   Temperature_C=NULL,
   DO = WC$Dissolved_Oxygen_Saturation,
   Dissolved_Oxygen_Saturation=NULL
  )
```

## c.     Write code to save the cleaned, revised dataset in your Data folder with a clear name indicating that it is the version to use in future analysis.

```{r}
write.csv(WC,here("data", "CAPDAP_WC_4Analysis.csv"),row.names=FALSE)
WC2<-read.csv(here("data", "CAPDAP_WC_4Analysis.csv"))
```

i will use "CAPDAP_WC_4Analysis.csv" in my future analysis

## d.     Perform exploratory data analysis as we discussed in class for the data in your project (hint: summary statistics, plots…)

```{r}
summary(WC2)
```

returns basic summary, but not intuitively informative, let's see the plots

```{r}
WC2$Site_Type<- as.factor(WC$Site_Type)
WC2$Date<-mdy(WC$Date)
WC2$Time<-hms(WC$Time)
WC$DateTime <- with(WC, Date + Time) 


ggplot(WC2,
  aes(x=Time, y=DO))+
    geom_point()+
    scale_x_time()+
    theme_bw()
```

I'm not sure exactly how to analyze a time analysis, but It looks like there are a series of DO values at 0 occuring on about every hour, but I anticipate that my analysis will change once the time scale adjusts, and I will look for more patterns during day vs night with that plotted.

```{r}
ggplot(WC,
  aes(x=pH, y=Chlorophyll))+
    geom_point()+
    theme_bw()
```

test for Erika

```{r}
WC<-WC%>%
    mutate(
      Site_Num = NULL,
      Site_Type=NULL,
      Date=NULL,
      Temperature_C=NULL,
      pH=NULL,
      Conductivity=NULL,
      Dissolved_Oxygen_Volume=NULL,
      Chlorophyll=NULL,
      MDY_HMS=NULL,
      timealtered = hms(WC$Time))%>%
        head(WC,n=15)
    

write.csv(WC,here("data", "Time_EG"),row.names=FALSE)
WC3<-read.csv(here("data", "Time_EG"))
```
