---
title: "Preliminary_Data_Analysis-LinearRegression"
author: "ML"
format: html
editor: visual
---

# Data Analysis: Linear Regression

CAPDAP Final Project 12/18/2024

In this document, I will explore the relationship between chlorophyll and temperature from the data file "CAPDAP_WC_4Analysis.csv."

I will use the following workflow model: plot data (and guess the results) → build the statistical model → Check the model assumptions → Interpret the model → Make a final plot

starting with a clean environment and package installation

```{r}
rm (list = ls())
library(tidyverse)
library(here)
library(ggfortify)
library(multcomp)
```

next, I'll read in the data

```{r}
WC<- read.csv(here("data", "CAPDAP_WC_4Analysis.csv"))
```

then I'll alter necessary data types

```{r}
WC$Site_Type <- as.factor(WC$Site_Type)
WC$Date<-ymd(WC$Date)
WC$Time<-hms(WC$Time)
```

My null hypothesis is that there is no relationship between chlorophyll and temperature in my Water Chestnut data.

My alternative hypothesis is that chlorophyll levels are positively correlated with water temperature. Warmer water temperatures are indicative of recent sunlight which suggests that more photosynthesis is processed in the river, resulting in more chlorophyll.

## Plotting Data

To test this, I'll first plot this proposed relationship (two continuous variables) using a scatterplot, and will later run a linear regression.

```{r}
ggplot(WC,aes(x=Temperature,y=Chlorophyll))+
  geom_point()+
  theme_bw()
```

There appears to be a strong positive relationship, indicating that as temperature increases, so does chlorophyll.

## Guessing Relationships

(The rise values can be mapped as (29.3-23) and the run as (5.2-1.8). M=6.3/3.4 or 1.85, so I estimate the slope to be \~1.85 and the y intercept to be around 19.5.

## Creating a Model

Next, I'll build my statistical model. Since I'm going to run a linear regression, I'll need to start with a model

```{r}
WCmodel <- lm(Temperature~Chlorophyll, data=WC)
WCmodel
```

So, here the slope is 1.733 and the y-int is 19.728. My guesses weren't too far off.

## Checking Assumptions

I'll need to check my assumptions before running any tests

```{r}
autoplot(WCmodel)
```

Reviewing these charts, I would conclude:

RvFit model - a little wave, but overall a linear model is likely appropriate

Normal Q-Q - looks good, pretty normal data fit to this line

Scale Location - this is a bit concerning, it indicates that there is an unevenness in the residuals

RvLeverage - There may be a few points with larger leverage here.

Overall, a polynomial model may fit this scenario better, but for now, I'm going to proceed with the linear model

## Running/Interpreting the Model

I can move onto my tests

```{r}
anova(WCmodel)
```

Here I see that there is a significant association between temperature and chlorphyll.

Let's dig in further.

```{r}
summary(WCmodel)
```

From this I see that the y-intercept is 19.7281 and the slope is 1.7325. The adjusted R2 is 0.3769, which means that about 37.69% of the variation in chlorophyll is associated with temperature (pretty good for a field test).

## Replotting the Data

Let's get a nicer looking plot

```{r}
ggplot(WC, aes(Temperature, Chlorophyll))+
  geom_point()+
  geom_smooth(method = "lm")+
  ggtitle("Comparing Temp and Chl in Water Chestnut")+
  labs(x = "Water Temperature in Degrees Celsius", y = "Chlorophyll (μg/L)")+
  theme_bw()+
  theme(
    axis.title.x=element_text(colour="blue", size=rel(1)), # adjust colour and size of x-axis title
    axis.title.y=element_text(colour="darkgreen", size=rel(1)), # adjust colour and size of x-axis title
      )
```

## Results Statement

There is a statistically significant and positive association between temperature and chlorophyll in the Oswegatchie River (F~1, 295~ = 180.01, p = \<\<\< 0.001, R^2^ = 0.3769).

This makes sense! When it is cold, the plant machinery slows down, resulting in less chlorophyll in the water.
